Negative Selection in the 20th Century

### 2020-06-16 why-negative-selection.Rmd
work and notes on census weighting. Some simplifications in the text.

Natural selection is taking place on most polygenic scores, and is broadly
"negative" in that characteristics that most people would consider undesirable
are being selected for. Effect sizes are larger for number of siblings than for
number of children. This is not because the strength of natural selection has
changed over time: effect sizes do not differ between early- and late-born
respondents, within either children or siblings. Instead, the difference is
probably due to ascertainment bias in UK Biobank. Indeed, natural selection is
concentrated among population subgroups that are under-represented in the
sample, including less-educated people and poorer households. When we weight 
the sample to correct for ascertainment bias, many effect sizes are considerably 
larger.

Among mothers who were older at first birth, and among parents with fewer sexual
partners, effect signs are reversed so that natural selection is broadly
"positive". The effects of PGS can be decomposed into two channels: an effect on
age at first live birth, and an opposite-signed effect on number of children
conditional on age at first live birth. This pattern may be explained by
economic theories of fertility. In these, higher potential earnings have two
opposite effects on fertility: a fertility-increasing income effect (higher
income makes children more affordable), and a fertility-lowering substitution
effect (time spent on childrearing has a higher cost in foregone earnings).
Among couples who can take advantage of coordination in childcare activities,
the income effect dominates; among single parents or those in unstable
relationships, the substitution effect dominates.

@@

* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1. 
* Need to make the weighting very watertight. 
  - Include income? 
  - Interact age and other variables?
  - Investigate "MRP" (Gelman 2005)
  - Understand reweighting on subsets of survey data.
  - Check for existing work
* Reweight other regressions (e.g. the omnibus regression in the appendix)
* Could be useful: f.2139 - age first had sex (includes "never had sex" 
  which may explain some of the many NAs for f.2141, num sex partners)
*  Need f.6138 to correctly calculate `age_fulltime_edu` including
  those who went to college.
* Look at geography, esp. in siblings regressions.
* control for age in n_partners regressions.

@@

res_combined %>% 
      mutate(
        score_name = fct_reorder(score_name, estimate, order_abs(2)),
        Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for age FLB")
      ) %>% 
      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
        color = p.value < 0.05/{{n_regs}})) + 
        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
        pval_geom_point() +
        pval_scale() +
        scale_fill_brewer(type = "qual", palette = 3) +
        my_labs() +
        my_theme()
```

>>>>

Natural selection is taking place on most polygenic scores, and is broadly
"negative" in that characteristics that most people would consider undesirable
are being selected for. Effect sizes are larger for number of siblings than for
number of children. This is not because the strength of natural selection has
changed over time. Instead, the difference is probably due to ascertainment bias
in UK Biobank. Indeed, natural selection is concentrated among population
subgroups that are under-represented in the sample, including less-educated
people and poorer households. When we weight the sample to correct for
ascertainment bias, many effect sizes are considerably larger.

Among mothers who were older at first birth, and among parents with fewer sexual
partners, effect signs are reversed. The effects of PGS can be decomposed into
two channels: an effect on age at first live birth, and an opposite-signed
effect on number of children conditional on age at first live birth. This
pattern may be explained by economic theories of fertility. In these, higher
potential earnings have two opposite effects on fertility: a
fertility-increasing income effect (higher income makes children more
affordable), and a fertility-lowering substitution effect (time spent on
childrearing has a higher cost in foregone earnings). Among couples who can take
advantage of coordination in childcare activities, the income effect dominates;
among single parents, or those in unstable relationships, the substitution effect
dominates.

@@

* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1. 
* Need to make the weighting very watertight. 
  - Include income? 
  - Interact age and other variables?
  - Investigate "MRP" (Gelman 2005)
  - Understand reweighting on subsets of survey data.
  - Check for existing work
  - Chris Hanretty suggested using 2011 census. Household size/type would also
    be good.
* Reweight other regressions (e.g. the omnibus regression in the appendix)
* Could be useful: f.2139 - age first had sex (includes "never had sex" 
  which may explain some of the many NAs for f.2141, num sex partners)
*  Need f.6138 to correctly calculate `age_fulltime_edu` including
  those who went to college.
* Look at geography, esp. in siblings regressions.
* control for age in n_partners regressions.

@@

res_combined %>% 
      mutate(
        score_name = pretty_names(score_name),
        score_name = fct_reorder(score_name, estimate, order_abs(2)),
        Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for age FLB")
      ) %>% 
      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
        color = p.value < 0.05/{{n_regs}})) + 
        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
        pval_geom_point() +
        pval_scale() +
        scale_fill_brewer(type = "qual", palette = 3) +
        my_labs() +
        my_theme()
```
